Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-07-30 elsarticle/elsarticle.tex
small corrections1

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we show that the population acts like a coherent organism under growth/depletion. 
Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with the Gompertz curve by augmenting the logistic growth equation and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. We thus find a novel parametrization bridging logistic and Gompertzian growth.
\end{abstract}

@@

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same time frame. While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model. We will then present a novel bridge between the these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

\section{Simple communicable disease models}
We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

A line of reasoning employed by Rypdal and Rypdal \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
\begin{align}
\label{eq:linSIR}
\frac{dY}{dt}& = \beta I\\
\frac{dI}{dt}& = (\beta - \alpha) I = \alpha (\phi(t) - 1) I,
\end{align}
and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.

@@

 At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
 The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
 One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
 Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate. In fact, Wang et al. \cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.

 Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
 Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. Estrada and Bartesaghi \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
 Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
 We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

\subsection{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richard's model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. We are thus prompted to look for another model which can explain the observed mortality patterns.

@@

Inspired by De Lauro et al. \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
\begin{equation}
\label{eq:microscopic}
dZ(t)= -\beta Z(t) dt + \sqrt{\sigma}dW(t),
\end{equation}
where $dW(t)$ is a delta-correlated Wiener process with zero mean, making $Z(t)$ a stochastic process. The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.

This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian. It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}. As argued by De Lauro et al. \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation. This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

@@

One could add network interaction necessitating a corresponding matrix version of Eq. \ref{eq:compareLog}
\begin{equation}
\label{eq:networkSIR}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j \quad \forall i,
\end{equation}
using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth or infection rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice that there is an implied causality from the infected to the susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{M}}{\partial x_i \partial x_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi \cite{estrada2022networked} provide further analysis on this topic.

@@

Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - M(t))} = -\beta$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - M(t) = \exp{(-\beta t)}$ \citep{shklovskii2005simple}. 
In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
Molski and Konarski \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

@@

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. Putting this all together, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} -\beta \ln{x(t)}.
 \end{equation}
 Note that for $N=1$, this becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term goes to zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}).

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, related although not identical to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

>>>>

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with expected initial logistic growth curves as predicted by traditional Susceptible-Infected-Recovered (SIR) models, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we connect logistic growth with Gompertzian growth by augmenting the logistic growth equation with higher order interaction terms, and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. Crucially, this augementation must be accompanied by causality-reversal where the source of growth is not the pool of infected but the pool of susceptible people. We thus find a novel bridge between logistic and Gompertzian growth, separate from the existing Richards model (also called $\theta$-logistic growth).
\end{abstract}

@@

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same timeframe. While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

\subsection{Simple communicable disease models}
We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

A line of reasoning employed by Rypdal and Rypdal (2020) \cite{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, are much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
\begin{align}
\label{eq:linSIR}
\frac{dY}{dt}& = \beta I\\
\frac{dI}{dt}& = (\beta - \alpha) I = \alpha (\phi(t) - 1) I,
\end{align}
and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.

@@

 At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
 The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
 One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
 Rypdal and Rypdal (2020) \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate. In fact, Wang et al. (2012) \cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.

 Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
 Herrmann and Schwartz (2020) \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. Estrada and Bartesaghi (2022) \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
 Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
 We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

\subsection{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richard's model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. (2020) \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

@@

The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. (2002) \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. We are thus prompted to look for another model which can explain the observed mortality patterns.

@@

Inspired by De Lauro et al. (2014) \cite{de2014stochastic}, the stressor can be modeled as a multiplicative stochastic dampening term along with a countering force of the immune system. 
As with all multiplicative processes, it is convenient to log-transform mortality and work in a dimensionless space such that $F(t)=Y(t)/Y_\infty$ and $Z(t)=\ln{F}(t)$, from which a natural perturbation model emerges,
\begin{equation}
\label{eq:microscopic}
dZ(t)= -\beta Z(t) dt + \sqrt{\sigma}dW(t),
\end{equation}
where $dW(t)$ is a delta-correlated Wiener process with zero mean, making $Z(t)$ a stochastic process. The first term on the right hand side represents the growth due to the stressor, while the second represents the stress response.

This perturbation model is called an Ornstein–Uhlenbeck process \citep{risken1996fokker}, where $Z(t)$ is Gaussian. It is also interpreted as Newton's equation of motion with friction and a random force (Langevin's equation), and a continuous version of an Auto-Regressive(1) model \citep{akaike1970statistical}. As argued by De Lauro et al. (2014) \cite{de2014stochastic}, the diffusion coefficient, $\sigma$, represents the strength of the perturbation. This is seen if we recast our perturbation model in terms of mortality while introducing a new parameter $C=\exp{(-\frac{\sigma}{2\beta})}$, viz. 
\begin{equation}
\label{eq:microscopicOriginal}
dF(t) = \left\{\frac{\sigma}{2}F(t) - \beta F(t)\ln\left[\frac{F(t)}{C}\right]\right\}dt + \sqrt{\sigma}F(t)dW(t).
\end{equation}

@@

One could add network interaction necessitating a corresponding matrix version of Eq. \ref{eq:compareLog}
\begin{equation}
\label{eq:networkSIR}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j \quad \forall i,
\end{equation}
using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth or infection rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice that there is an implied causality from the infected to the susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{M}}{\partial x_i \partial x_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. Estrada and Bartesaghi (2022) \cite{estrada2022networked} provide further analysis on this topic.

@@

Under this model, correlation between entities is present at all orders in the original domain and all the nodes in the network communicate instantaneously.

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

@@

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. Putting this all together, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)}.
 \end{equation}
 Note that for $N=1$, this becomes the standard logistic function, while higher order terms will move closer and closer to the Gompertz curve as the hypergerometric term goes to zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}).

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

### 2022-07-30 letterToEditor/letter.tex
small corrections1

If you do decide to review this paper you may find the enclosed list of possible referees helpful. 
Some of them are listed in the references of the manuscript.

@@

\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
Dipartimento di Matematica and TIRES, University of Bari (Ret).
Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model. We use their concept of growth rooted in stochastic phenomena.
\item Salvatore De Martino,
demartino247@gmail.com, 
Dipartimento di Ingegneria dell’Informazione ed Elettrica.
Same as above
\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
Has written on the SIR model and chaotic dynamics.
\item Gabor Vattay, vattay@elte.hu, Eötvös Loránd University Budapest, physics, SIR models
Has written a paper on SIR models showing that there are ``initial transients'' in the observations which are explained by us in the Gompertz model.
\item E. B. Postnikov, postnikov@kursksu.ru, Kursk State Univesity, physics, SIR models.
Has done work on SIR models using a mean field approach, but has a broach research interest that could be a benefit when reviewing our paper.
\item L. Kocarev, lkocarev@manu.edu.mk, Macedonian Academy of Sciences and Arts
, physics and mathematics, SIR models.
Has written papers on SIR models as they relate to non-markovian models.
\item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria. Has done extensive work on long-range interaction and statistical mechanics. He might be able to see connections with his famous Tsallis entropy.

>>>>

If you do decide to review this paper you may find the enclosed list of possible referees helpful. Some of them are listed in the references of the manuscript. Due to the conclusions of our paper, I request that this paper receives more reviews than your average, and at least 3 reviews.

@@

\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
Dipartimento di Matematica and TIRES, University of Bari (Ret).
Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model. We use their concept of growth rooted in stochastic phenomena.
\item Salvatore De Martino,
demartino247@gmail.com, 
Dipartimento di Ingegneria dell’Informazione ed Elettrica.
Same as above
\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
Has written on the SIR model and chaotic dynamics.
\item Constantino Tsallis, tsallis@cbpf.br, Complexity Science Hub Vienna, Austria. Has done extensive work on long-range interaction and statistical mechanics. He might be able to see connections with his famous Tsallis entropy.
